---
title: Homework Set 7
author: Ansel George
output: pdf_document
---

```{r,message=F}
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(10)
```

# Part I

## 1
*Implement a program in R to simulate the discrete Wright-Fisher model under
genic selection (without mutation). To this end, start with a given initial
frequency $p_0$. Given that the population allele-frequency in generation $k$
is $p_k$, the frequency in the next generation $k + 1$ is given by*

\begin{align}
p_{k+1} = \frac{X_{k+1}}{2N}
\end{align}

*where $X_{k+1}$ is distributed according to a Binomial distribution with $2N$
trials and success probability $\Psi_k$ given by the selection-adjusted
population allele frequency, as discussed in class. This dynamics is repeated
for several generations, until the frequency is either equal to $0$ (the
beneficial allele is lost) or equal to $1$ (the beneficial allele fixed). It is
advisable to stop the simulation after a large number of generations (for
example $20000$), even if the allele is not yet absorbed.*

```{r}
getNextGenSelection <- function(p, popsize, s) {
  psi <- (1+s)*p/(s*p+1) # adjust p by selection coefficent
  return(rbinom(length(p), popsize, psi)/popsize)
}

makeWFSelection <- function(p, s, ngenerations, popsize, reps) {
  pop <- matrix(0, nrow=reps, ncol=ngenerations)
  pop[,1] <- rep(p, reps)

  for (i in 2:ngenerations) {
    pop[,i] <- getNextGenSelection(pop[,i-1], popsize, s)
  }
  return(pop)
}
```


## 2
*Using this implementation, simulate the Wright-Fisher model $1000$ times with
$N = 2000$, initial frequency $0.2$, and selection coefficient $s_1 = 0.0005$
and $s_2 = 0.001$, respectively. For each selection coefficient, what is the
fraction of simulations where the beneficial allele gets fixed.*

```{r}
N <- 2000
reps <- 1000
ngenerations <- 20000
s1 <- .0005
s2 <- .001
p <- .2
```

```{r}
# simulate populations
wfpop1 <- makeWFSelection(p, s1, ngenerations, 2*N, reps)
wfpop2 <- makeWFSelection(p, s2, ngenerations, 2*N, reps)
```

```{r}
mean(wfpop1[,ngenerations] == 1)
mean(wfpop2[,ngenerations] == 1)
```


## 3
*Plot the fixation probability discussed in class*

\begin{align}
\mathbb{P}_x \{T_1 < T_0\} = \frac{1-e^{-2\alpha x}}{1-e^{-2\alpha}}
\end{align}

*as a function of the initial frequency $x \in [0, 1]$ for $\alpha_1 = 2N_{s1}$
and $\alpha_2 = 2N_{s2}$. In addition, plot the fixation probability under
neutrality $P_x \{T_1 < T_0 \} = x$. Are the values for the fixation
probability estimated from the simulations compatible with these plots?*

```{r}
fixationProb <- function(x, a) {
  return( (1-exp(-2*a*x))/(1-exp(-2*a)) )
}
```

```{r}
N <- 2000
s1 <- .0005
s2 <- .001

a1 <- 2*N*s1
a2 <- 2*N*s2
p <- seq(0,1,.001)
```

```{r}
qplot(p, fixationProb(p, a1))
qplot(p, fixationProb(p, a2))
qplot(p, p)
```

```{r}
fixationProb(.2, a1)
fixationProb(.2, a2)
```


# Part II

*Consider the discrete Wright-Fisher model for a population of size $N$ ($2N$
haploids) with recurrent mutation (no selection), where $u$ is the per
generation probability of $A \rightarrow a$ mutations and $v$ is the
probability of $a \rightarrow A$ mutations*

## 1
*We had seen in class, that the stationary distribution of the population-allele
frequency $y$ of A alleles for the corresponding Wright-Fisher diffusion is given
by the Beta distribution*

\begin{align}
\pi_{\beta_1,\beta_2}(y) = \frac{1}{B(\beta_1,\beta_2)} y^{\beta_2-1}(1-y)^{\beta_1-1}
\end{align}

*with $\beta_1 = 4Nu$ and $\beta_2 = 4Nv$. Plot this distribution using the
R-function `dbeta` with $N = 2500$ for $u = v = 0.0002$, $u = v = 0.0001$, and $u
= v = 0.00005$. Set the limits of the y-axis as $[0,4]$.*

```{r}
N <- 2500
u1 <- .0002
v1 <- .0002
u2 <- .0001
v2 <- .0001
u3 <- .00005
v3 <- .00005

p <- seq(0,1,.001)
```

```{r}
a1 <- 4*N*u1
b1 <- 4*N*v1
qplot(p, dbeta(p, a1, b1), ylim=c(0,4))

a2 <- 4*N*u2
b2 <- 4*N*v2
qplot(p, dbeta(p, a2, b2), ylim=c(0,4))

a3 <- 4*N*u3
b3 <- 4*N*v3
qplot(p, dbeta(p, a3, b3), ylim=c(0,4))
```


## 2
*Implement a program in R to simulate the discrete Wright-Fisher model with
recurrent mutation (no selection). Again, start with a given initial frequency
$p_0$. Given that the population allele-frequency in generation $k$, the
frequency in the next generation is given by*

\begin{align}
p_{k+1} = \frac{X_{k+1}}{2N}
\end{align}

*where $X_{k+1}$ is distributed according to a Binomial distribution with $2N$
trials and success probability $\Psi_k$ given by the mutation-adjusted
population allele frequency, as discussed in class. For mutation probabilities
$u = v = 0.0002$, $u = v = 0.0001$, and $u = v = 0.00005$ perform $1000$
simulations each. However, for each replicate, draw the initial frequency $p_0$
from the respective stationary distribution (R-function `rbeta`). For each set of
mutation probabilities, plot a histogram of the initial frequencies, and a
histogram of the frequencies after $500$ generations (use `breaks=50` for the
histograms). Compare the histograms of the initial frequencies with the
histogram of the frequencies after $500$ generations, and with the plot of the
stationary distribution.*

```{r}
getNextGenMutation <- function(p, popsize, u, v) {
  psi <- p - p*u + (1-p)*v # adjust p by mutation rate
  return(rbinom(length(p), popsize, psi)/popsize)
}

makeWFMutation <- function(p, u, v, ngenerations, popsize, reps) {
  pop <- matrix(0, nrow=reps, ncol=ngenerations)
  a <- 4*popsize*u
  b <- 4*popsize*v
  pop[,1] <- rbeta(reps, a, b)

  for (i in 2:ngenerations) {
    pop[,i] <- getNextGenMutation(pop[,i-1], popsize, u, v)
  }
  return(pop)
}
```

```{r}
ngenerations <- 500
reps <- 1000

wf_m1 <- makeWFMutation(p, u1, v1, ngenerations, N, reps)
wf_m2 <- makeWFMutation(p, u2, v2, ngenerations, N, reps)
wf_m3 <- makeWFMutation(p, u3, v3, ngenerations, N, reps)
```

```{r}
header <- paste('T', seq(1,ngenerations), sep='')

colnames(wf_m1) <- header
wf_m1 <- tbl_df(wf_m1)

colnames(wf_m2) <- header
wf_m2 <- tbl_df(wf_m2)

colnames(wf_m3) <- header
wf_m3 <- tbl_df(wf_m3)
```

```{r}
wf_m1 %>% select(T1, T500) %>%
  gather(Generation, p) %>%
  ggplot() + aes(x=p) + facet_grid(.~Generation) + geom_histogram(bins=50)

wf_m1 %>% select(T1, T500) %>%
  ggplot() + aes(x=sort(T1), y=sort(T500)) + geom_point() +
  geom_abline(intercept=0, slope=1) + xlab("T1") + ylab("T500")

wf_m2 %>% select(T1, T500) %>%
  gather(Generation, p) %>%
  ggplot() + aes(x=p) + facet_grid(.~Generation) + geom_histogram(bins=50)

wf_m2 %>% select(T1, T500) %>%
  ggplot() + aes(x=sort(T1), y=sort(T500)) + geom_point() +
  geom_abline(intercept=0, slope=1) + xlab("T1") + ylab("T500")

wf_m3 %>% select(T1, T500) %>%
  gather(Generation, p) %>%
  ggplot() + aes(x=p) + facet_grid(.~Generation) + geom_histogram(bins=50)

wf_m3 %>% select(T1, T500) %>%
  ggplot() + aes(x=sort(T1), y=sort(T500)) + geom_point() +
  geom_abline(intercept=0, slope=1) + xlab("T1") + ylab("T500")
```


# Part III

*For a given population allele frequency $y$, the heterozygosity*

\begin{equation}
h(y) := 2y(1-y)
\end{equation}

*yields the probability that $2$ randomly paired gametes do not carry the same
allele. With initial frequency $x$, how does the expected heterozygosity*

\begin{equation}
\mathbb{E}_x [h(Y_t)]
\end{equation}

*change over time under the neutral Wright-Fisher diffusion with just genetic
drift (no mutation, no selection). Thus, we have $\mu(y) = 0$ and $\sigma^2 (y)
= y(1 - y)$. To this end, consider the derivative with respect to time*

\begin{align}
\frac{\partial}{\partial t} \mathbb{E}_x [h(Y_t)] &= \frac{\partial}{\partial t} \int_0^1 h(y)p(t;x,y)dy \\
  &= \int_0^1 h(y) \frac{\partial}{\partial t}\{p(t;x,y)\}dy
\end{align}

*which follows directly from the definition of the expected value. Compute this
derivative, by substituting the forward equation and using integration by
parts. You can assume that the terms for $y = 0$ and $y = 1$ are $0$. What does
this derivative imply for the change of the expected heterozygosity over time?*


\begin{align}
\frac{\partial}{\partial t} \mathbb{E}_x [h(Y_t)] &= \int_0^1 h(y) \frac{\partial}{\partial t}\{p(t;x,y)\}dy \\
  &= \int_0^1 h(y) \Big( \frac{\partial}{\partial y} \{ \mu (y) p(t;x,y) \} + \frac{1}{2} \frac{\partial^2}{\partial y^2} \{ \sigma^2 (y) p(t;x,y) \} \Big)dy \\
  &= \int_0^1 h(y) \Big( \frac{\partial}{\partial y} \{ 0 p(t;x,y) \} + \frac{1}{2} \frac{\partial^2}{\partial y^2} \{ \sigma^2 (y) p(t;x,y) \} \Big)dy \\
  &= \int_0^1 h(y) \frac{1}{2} \frac{\partial^2}{\partial y^2} \{ \sigma^2 (y) p(t;x,y) \} dy \\
  &= \big[\frac{1}{2} h(t) \frac{\partial}{\partial y} \{ \sigma^2 (y) p(t;x,y) \} \big]^1_0 - \int_0^1 h^\prime (y) \frac{1}{2} \frac{\partial}{\partial y} \{ \sigma^2 (y) p(t;x,y) \} dy \\
  &= 0 - \int_0^1 h^\prime (y) \frac{1}{2} \frac{\partial}{\partial y} \{ \sigma^2 (y) p(t;x,y) \} dy \\
  &= -\big[\frac{1}{2} h^\prime (t) \sigma^2 (y) p(t;x,y) \big]^1_0 + \int_0^1 h^{\prime\prime}(y) \frac{1}{2}\sigma^2 (y) p(t;x,y)dy \\
  &= 0 + \int_0^1 h^{\prime\prime}(y) \frac{1}{2}\sigma^2 (y) p(t;x,y)dy \\
  &= \int_0^1 (-4) \frac{1}{2}(y(1-y)) p(t;x,y)dy \\
  &= \int_0^1 -2(y(1-y) p(t;x,y)dy \\
  &= - \int_0^1 2(y(1-y) p(t;x,y)dy \\
  &= - \int_0^1 h(y) p(t;x,y)dy \\
  &= -\mathbb{E}_x [h(Y_t)] \\
\implies \frac{d \mathbb{E}_x [h(Y_t)]}{\mathbb{E}_x [h(Y_t)]} &= -dt \\
\implies ln\big(\mathbb{E}_x [h(Y_t)]\big) &= -t + ln(C) \\
\implies \mathbb{E}_x [h(Y_t)] &= Ce^{-t} \\
\implies \mathbb{E}_x [h(Y_t)] &= Y_0 e^{-t}\textrm{, given }\mathbb{E}_x [h(Y_{t=0})] = Y_0
\end{align}

This result means that the expected heterozygosity in a population undergoing
Wright-Fisher diffusion will decline exponentially over time.
