---
title: "Homework 5"
author: "Ansel George"
output: pdf_document
---

```{r}
library(ggplot2)

set.seed(10)
```


# 1

*Consider the rescaled relative population size function*

\[
  f(t) = \left\{
     \begin{array}{@{}l@{\thinspace}l}
       1  & \text{ for } t < 0.05 \\
       10 & \text{ for } t \ge 0.05 \\
     \end{array}
   \right.
\]

*which models a recent bottleneck extending up to the present.*

## a)

*Assuming $N_0 = 10000$, what is the size of the population at present? What was
the size before the bottleneck? How many generations ago did the bottleneck
start?*

$N_0$, the reference population size, is defined as the population size at
present, which in this case is $10000$.

The population size at the bottleneck, which occurred at time $t = .05$,
evaluates to $10 * 10000 = 100000$.

The time of the bottleneck (in the number of generations) is $2 N_0 t = 2 *
10000 * .05 = 1000$.


## b)

*Compute the cumulative coalescent rate function:*

\begin{align*}
\Lambda(t) := \int_{s=0}^t \frac{1}{f(s)} ds
\end{align*}

When $t < .05$
\begin{align*}
\Lambda(t) &= \int^t_{s=0} \frac{1}{1} ds = t
\end{align*}

When $t \ge .05$
\begin{align*}
\Lambda(t) &= \int^{.05}_{s=0} \frac{1}{1} ds + \int^{t}_{s=0.05} \frac{1}{10} ds \\
  &= .05 + \frac{t}{10} - \frac{.05}{10} \\
  &= \frac{t}{10} + .045
\end{align*}

Together:

\[
  \Lambda(t) = \left\{
     \begin{array}{@{}l@{\thinspace}l}
       t  & \text{ for } t < 0.05 \\
       \frac{t}{10} + .045 & \text{ for } t \ge 0.05
     \end{array}
   \right.
\]


## c)

*In class we showed that*

\begin{align*}
\mathbb{P} \big\{ T^{(k)}_k > t \big\} = \mathrm{e}^{- { k \choose 2 } \Lambda(t)}
\end{align*}

*holds for the first-coalescence time $T_k^{(k)}$ of $k$ lineages. Compute the
expected value*

\begin{equation*}
\mathbb{E}\big[ T_k^{(k)} \big] = \int_{t=0}^\infty \mathbb{P}\big\{T^{(k)}_k > t \big\}dt
\end{equation*}

Separating the equation into two conditions, when $t < .05$:

\begin{align*}
\mathbb{E}\big[ T_k^{(k)} \big] &= \int_{t=0}^\infty \mathbb{P}\big\{T^{(k)}_k > t \big\}dt \\
  &= \int_{t=0}^\infty \mathrm{e}^{- { k \choose 2 } \Lambda(t)} dt \\
  &= \int_{t=0}^{.05} \mathrm{e}^{- { k \choose 2 } t} dt + \int_{t=0.05}^\infty \mathrm{e}^{- { k \choose 2 } \big( \frac{t}{10} + .045 \big) } dt\\
  &= \frac{-1}{ { k \choose 2 } } \big( \left. e^{ -{ k \choose 2 } t } \right|^{.05}_0 \big) + \frac{-10}{ { k \choose 2 } } e^{ -{ k \choose 2 } .045 } \big( \left.  e^{ -{ k \choose 2 } \frac{t}{10} } \right|^{\infty}_{.05} \big) \\
  &= \frac{1}{ { k \choose 2 } } \Big( 1 - e^{ -{ k \choose 2 } .05 } + 10e^{ -{ k \choose 2 } .045 } e^{ -{ k \choose 2 } \frac{.05}{10} } \Big) \\
  &= \frac{1}{ { k \choose 2 } } \Big( 1 - e^{ -{ k \choose 2 } .05 } + 10e^{ -{ k \choose 2 } .05 } \Big) \\
  &= \frac{1}{ { k \choose 2 } } \Big( 1 + 9e^{ -{ k \choose 2 } .05 } \Big)
\end{align*}


# 2

*Consider a sample of haplotypes observed at 3 loci. The possible alleles at the
first locus are $\{a, A\}$, at the second locus $\{b, B\}$, and at the third
locus $\{c, C\}$. Denote by $n_{a,b,c}$ the number of haplotypes in the sample
with alleles $(a, b, c)$ at the respective loci, $n_{a,b,C}$ the number of
haplotypes with alleles $(a, b, C)$, and so forth... Assume that we observe a
sample of haplotypes of size 20 given by:*

\begin{equation}
  \left(
  \begin{array}{c}
    n_{a,b,c} \\
    n_{a,b,C} \\
    n_{a,B,c} \\
    n_{a,B,C} \\
    n_{A,b,c} \\
    n_{A,b,C} \\
    n_{A,B,c} \\
    n_{A,B,C} \\
  \end{array}
  \right) =
  \left(
    \begin{array}{c}
      2 \\
      4 \\
      2 \\
      4 \\
      3 \\
      1 \\
      2 \\
      2 \\
    \end{array}
   \right)
\end{equation}


*We had defined linkage disequilibrium $r^2_{a,c}$ as*

\begin{equation*}
r^2_{a,c} := \frac{D^2_{a,c}}{\hat{p}_{a}\hat{p}_{A}\hat{p}_{c}\hat{p}_{C}}
\end{equation*}

*with*

\begin{equation*}
D_{a,c} := \hat{p}_{a,c} - \hat{p}_{a}\hat{p}_{c}
\end{equation*}

*with empirical haplotype frequency*

\begin{equation*}
\hat{p}_{a,c} = \frac{n_{a,c}}{n}
\end{equation*}

*and the empirical allele frequency*

\begin{equation*}
\hat{p}_{a} = \frac{n_{a}}{n}
\end{equation*}

*Here $n_{a,c}$ is the number of haplotypes in the sample that have allele $a$
at the first locus and allele $c$ at the third, and $n_a$ is the number of
haplotypes with allele $a$ at the first locus. The same definitions hold for
the other alleles. Compute $r_{a,b}$, $r_{a,c}$, and $r_{b,c}$ for the sample
given equation (1).*


$p_a = \frac{2+4+2+4}{20} = 0.6$  
$p_b = \frac{2+4+3+1}{20} = 0.5$  
$p_c = \frac{2+2+3+2}{20} = 0.45$  

$p_{a,b} = \frac{2+4}{20} = 0.3$  
$p_{a,c} = \frac{2+2}{20} = 0.2$  
$p_{b,c} = \frac{2+3}{20} = 0.25$  

$D_{a,b} = \hat{p}_{a,b} - \hat{p}_{a}\hat{p}_{b} = .3 - .6*0.5 = 0$  
$D_{a,c} = \hat{p}_{a,c} - \hat{p}_{a}\hat{p}_{c} = .2 - .6*0.45 = -0.07$  
$D_{b,c} = \hat{p}_{b,c} - \hat{p}_{b}\hat{p}_{c} = .25 - .5*0.45 = 0.025$  

$r^2_{a,b} = \frac{D^2_{a,b}}{\hat{p}_{a}\hat{p}_{A}\hat{p}_{b}\hat{p}_{B}} = \frac{(0)^2}{0.6(1-.06).5(1-.5)} = 0$  
$r^2_{a,c} = \frac{D^2_{a,c}}{\hat{p}_{a}\hat{p}_{A}\hat{p}_{c}\hat{p}_{C}} = \frac{(0.07)^2}{0.6(1-.06).45(1-.45)} = 0.08249158$  
$r^2_{b,c} = \frac{D^2_{b,c}}{\hat{p}_{b}\hat{p}_{B}\hat{p}_{c}\hat{p}_{C}} = \frac{(0.025)^2}{0.5(1-.05).45(1-.45)} = 0.01010101$  

$r_{a,b} = \sqrt{0} = 0$  
$r_{a,c} = \sqrt{0.08249158} = 0.2872135$  
$r_{b,c} = \sqrt{0.01010101} = 0.1005038$  


# 3

*Given a sample size $n$, a recombination rate $\rho$, and the sequence length
$L$, recall that in the full ancestral recombination graph (not reduced),
recombination occurs at rate $\frac{\rho}{2}(L - 1)$ along each lineage, and
the lineage branches into two new lineages. Furthermore, coalescence occurs
between each pair of lineages at rate one. Consider the so-called genealogical
ancestral process $G_n (t)$ that gives the total number of ancestral lineages
in the full ARG at time $t$ before present (regardless which loci these
lineages are ancestral to). This process starts out with $G_n (0) = n$
lineages. Whenever there are $k$ lineages, this number is increased to $k + 1$
at rate $k \frac{\rho}{2}L$ or decreased to $k - 1$ at rate ${k \choose 2}$.*

*Implement an R-function that simulates realizations of $G_n (t)$. To this end,
for each trajectory, set the initial number of lineages to $n = k$. After an
exponentially distributed time with parameter $n \frac{\rho}{2}(L - 1) + {n
\choose 2}$ the number of lineages increases by one with probability*

\begin{equation*}
\frac{k\frac{\rho}{2}(L-1)}{k\frac{\rho}{2}(L-1) + {k \choose 2}}
\end{equation*}

*and decreases by one with probability*

\begin{equation*}
\frac{ {k \choose 2} }{k\frac{\rho}{2}(L-1) + {k \choose 2}}
\end{equation*}

*Update the number of lineages $k$ and repeat this procedure with the updated
number until there is only one lineage left (the ultimate most recent common
ancestor). Use this program to simulate $1000$ trajectories with $n = 10$, $L =
200$, and $\rho = 0.02$. How many trajectories reach $1$ before $t = 20$?*

```{r}
simulateAncestralProcess <- function(n, L, rho) {
  k <- n
  time <- 0
  while(k != 1) {
    t <- rexp(1, (choose(k, 2) + k*rho*(L-1)/2))
    p <- choose(k, 2) / (choose(k, 2) + k*rho*(L-1)/2)
    if (runif(1) < p) {
      k <- k - 1
    } else {
      k <- k + 1
    }
    time <- time + t
  }
  return(time)
}
```

```{r}
n <- 10
L <- 200
rho <- .02
iter <- 1000

simulations <- list()
simulations <- rep(0, iter)
for (i in 1:iter) {
  simulations[i] <- simulateAncestralProcess(n, L, rho)
}

sum((simulations) < 20)
```

For this simulation, 918 lineages out of 1000 coalesce to a single common
ancestor in less than 20 generations.


# 4

*Consider two loci with possible alleles $\{a, A\}$ at the first locus and $\{b,
B\}$ at the second loci, and recombination rate $\rho$ between these two loci.
In class, we had shown that under the Coalescent with Recombination, the
expected value of the linkage disequilibrium*

\begin{equation*}
\mathbb{E}[r^2] = \mathbb{E}\bigg[ \frac{D_{AB}}{p_A p_a p_B p_b} \bigg]
\end{equation*}

*can be approximated by*

\begin{equation*}
\sigma^2_d = \frac{\rho + 10}{\rho^2 + 13\rho + 22}
\end{equation*}


## a)

*Using R, plot this function for $\rho$ between 0 and 50. What is the value of
the function at $\rho$ = 0? What is 5% of this value?*

```{r}
f <- function(x) {
  return((x+10)/(x^2 + 13*x + 22))
}

rho <- seq(0,50,.01)

qplot(rho, f(rho))
```

When $\rho=0$, $\sigma^2_d = \frac{10}{22}$. 5% of this value is $\frac{1}{44}$.


## b)

*Calculate the (positive) $\rho$ for which $\sigma_d^2$ is equal to 5% of its
value at $\rho = 0$. Thus, the correlation between two sites at this distance
$\rho$ is only 5%.*

The expression for $\sigma^2_d$ is 5% of the value at $\rho = 0$ when $\rho =
41.15638$

\begin{align}
.05 \frac{10}{22} &= \frac{\rho + 10}{\rho^2 + 13\rho + 22} \\
\frac{1}{44} &= \frac{\rho + 10}{\rho^2 + 13\rho + 22} \\
\rho^2 + 13\rho + 22 &= 44\rho + 440 \\
(\rho - \frac{31}{2})^2 &= 418 + \frac{31^2}{2^2} \\
\rho - \frac{31}{2} &= \sqrt{418 + \frac{31^2}{2^2}} \\
\rho &= \sqrt{418 + \frac{31^2}{4}} + \frac{31}{2} \\
  &\approx 41.15638
\end{align}


## c)

*Recall the formula $\rho = 4N_e r$, that relates the coalescent-scaled
recombination rate $\rho$ to the per generation recombination probability $r$.
Assuming $N_e = 10000$, a per base-pair per generation recombination
probability of $r = 1.25 * 10^{-8}$ , how many base-pairs (bp) does this $\rho$
correspond to?*

From the units of the expression, $\rho = 4N_e r$:

$r \textrm{ recombinations per bp } * x \textrm{ bp } = \textrm{ recombinations
per generation }$

\begin{align}
\rho &= r N_e r_{bp} x_{bp} \\
\implies x_{bp} &= \frac{\rho}{4 N_e r_{bp}} \\
  &= \frac{41.15638}{1.25 * 10^-8 * 4 * 10000} \\
  &= 82312.76
\end{align}
